---
title: "Estimating a flood inundation probability map"
author: "Yixuan"
date: "2024-03-30"
output: html_document
---

```{r setup, include=FALSE,message = FALSE,cache=TRUE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
library(knitr)
```

# 1. Set up


```{r libraries, warning = FALSE, message = FALSE}

#install.packages("dplyr")
#install.packages("sf")
#install.packages("ggplot2")
#install.packages("viridis")
#install.packages("gridExtra")
#install.packages("caret")
#install.packages("tidyr")
#install.packages("pROC")
#install.packages("plotROC")
#update.packages(ask = FALSE)

library(dplyr)
library(sf)
library(ggplot2)
library(viridis)
library(gridExtra)
library(caret)
library(tidyr)
library(pROC)
library(plotROC)


```


```{r mapTheme, echo=TRUE}
plotTheme <- function(base_size = 12) {
  theme(
    plot.title = element_text(size = base_size, face = "bold"),
    plot.subtitle = element_text(size = base_size * 0.83),
    plot.caption = element_text(size = base_size * 0.5),
    axis.title.x = element_blank(), 
    axis.title.y = element_text(size = base_size),
    axis.text.x = element_text(size = base_size * 0.83, hjust = 1),
    axis.text.y = element_text(size = base_size * 0.83),
    axis.ticks = element_blank(),
    panel.grid.major = element_line(colour = "#D0D0D0", size = 0.2),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.border = element_rect(colour = "darkgrey", fill = NA, size = 0.5), 
    legend.position = "right", 
    legend.background = element_blank(),
    legend.text = element_text(size = base_size * 0.67),
    legend.title = element_text(size = base_size-7)
  )
}


mapTheme <- function(base_size = 12, title_size = 16) {
  theme(
    # Basic text properties
    text = element_text(color = "black"),
    plot.background = element_rect(fill = "white", colour = NA), 
    panel.background = element_rect(fill = "white", colour = NA), 
    panel.border = element_rect(colour = "darkgrey", fill=NA, size=0.5), 
    
    # Title, subtitle, and caption customization
    plot.title = element_text(size = title_size, hjust = 0.5),
    plot.subtitle = element_text(face = 'italic', size = base_size, hjust = 0.5),
    plot.caption = element_text(hjust = 0),
    
    # Axis titles and text customization
    axis.title = element_text(size = base_size),
    axis.text = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    
    # Legend customization
    legend.background = element_blank(),
    legend.title = element_text(face = "italic"),
    legend.text = element_text(face = "italic",size = base_size * 0.67),
    legend.key.size = unit(.5, 'line'),
    
    # Panel grid customization
    panel.grid.major = element_line("grey100", size = 0.1),
    panel.grid.minor = element_blank(),
    
    # Strip customization
    strip.background = element_blank(),
    strip.text.x = element_text(size = 15)
  )
}


```

```{r input_boundary, warning = FALSE, message = FALSE, results = "hide"}
denverBoundary = st_read("D:/Upenn/23spring/CPLN 6750 Land Use and Environmental Modeling/Assignment3. Spatial flood inundation probability/Landuse_A3/GIS/1-GIS FILE/MyProject1/EXPORT/mygeodata/border_level6_polygon.geojson")
denverBoundary <- 
  st_read(file.path("D:/Upenn/23spring/CPLN 6750 Land Use and Environmental Modeling/Assignment3. Spatial flood inundation probability/Landuse_A3/GIS/1-GIS FILE/MyProject1/EXPORT/mygeodata/border_level6_polygon.geojson")) 

# Check if the geometry is valid
valid = st_is_valid(denverBoundary)
if (any(!valid)) {
  # Attempt to fix invalid geometries
  denverBoundary = st_make_valid(denverBoundary)
}

print(denverBoundary)
plot(denverBoundary)


```

# 2. Data Wraggling

```{r}
# Replace 'path/to/your/shapefile' with the actual path to your .shp file
calgary_fishnet <- st_read("D:/Upenn/23spring/CPLN 6750 Land Use and Environmental Modeling/Assignment3. Spatial flood inundation probability/Landuse_A3/GIS/1-GIS FILE/MyProject1/EXPORT/Calgary_Fishnet/Calgary_fishnet.shp")

calgary_fishnet <- calgary_fishnet %>%
  select(Mean, MAJORITY, MEAN_1, MAJORITY_1) %>%
  rename(dem = Mean,
         landcover = MAJORITY,
         distance_to_streams = MEAN_1,
         inundation = MAJORITY_1)

calgary_fishnet$landcover <- factor(calgary_fishnet$landcover)
calgary_fishnet$inundation<- factor(calgary_fishnet$inundation)

```

# 3. Maps of four original yet statistically significant features

### note: 1 we may export the maps from arcgis so we have propoer legend for each map.;2 the inundation should be replaced by slope; 3.also please remember to include the watershed feature

```{r}
calgary_fishnet.long <- 
  gather(calgary_fishnet, Variable, value, -geometry)
calgary_fishnet.long$value <- as.numeric(as.character(calgary_fishnet.long$value))

if (any(is.na(calgary_fishnet.long$value))) {
    stop("Conversion to numeric resulted in NA values, please check the data")
}
vars <- unique(calgary_fishnet.long$Variable)
mapList <- list()


for(i in vars){
  mapList[[i]] <- 
    ggplot() +
      geom_sf(data = filter(calgary_fishnet.long, Variable == i), aes(fill=value), colour=NA) +
      scale_fill_viridis(name="") +
      labs(title=i) +
      mapTheme()}

do.call(grid.arrange,c(mapList, ncol=2, top="Potential variables influencing innundation by Fishnet in Calgary"))

```

# 4. Model building

## 4.1. Partition training and test sets

```{r training_set}
set.seed(3456)
trainIndex <- createDataPartition(calgary_fishnet$inundation, p = .70,
                                  list = FALSE,
                                  times = 1)

calgaryTrain <- calgary_fishnet[ trainIndex,]
calgaryTest  <- calgary_fishnet[-trainIndex,]

```

## 4.2. Make a binomial model

```{r firstModel, warining = FALSE, message = FALSE}
Model <- glm(inundation ~ ., 
             family = binomial(link = "logit"), 
             data = calgaryTrain %>%
                      as.data.frame() %>%
                      select(-geometry))

summary(Model)

```

## 4.3. Model validation

```{r predict_first}

classProbs <- predict(Model, calgaryTest, type="response")

hist(classProbs)

```

```{r plot_preds}
testProbs <- data.frame(obs = calgaryTest$inundation,
                        pred = classProbs)

ggplot(testProbs, aes(x = pred, fill=as.factor(obs))) + 
  geom_density() +
  facet_grid(obs ~ .) + 
  xlab("Probability") +
  ylab("Frequency")+
  geom_vline(xintercept = .5) +
  scale_fill_manual(values = c("dark blue", "dark green"),
                      labels = c("No inundation","Inundation"),
                      name = "")+
  plotTheme()

```

### 4.3.1 Confusion metrics

```{r confusion_matrix, message = FALSE, warning = FALSE}
testProbs$predClass  = ifelse(testProbs$pred > .5 ,1,0)

caret::confusionMatrix(reference = as.factor(testProbs$obs), 
                       data = as.factor(testProbs$predClass), 
                       positive = "1")
```

### 4.3.2. ROC Curve

```{r roc_curve, message = FALSE, warning = FALSE}
library(ggplot2)
library(pROC)

testProbs$obs <-as.numeric(testProbs$obs)
testProbs$predClass <-as.numeric(testProbs$predClass)

ggplot(testProbs, aes(d = obs, m = pred)) + 
  geom_roc(n.cuts = 50, labels = FALSE) + 
  style_roc(theme = theme_grey) +
  geom_abline(slope = 1, intercept = 0, size = 1.5, color = 'grey') 

```

How about the area under the curve?

```{r auc, warning = FALSE}

auc(testProbs$obs, testProbs$pred)

```

### 4.3.3. Cross validation

```{r k_fold, warning = FALSE, message = FALSE}
ctrl <- trainControl(method = "cv", 
                     number = 100, 
                     p = 0.7, 
                     savePredictions = TRUE)

cvFit <- train(as.factor(inundation) ~ .,  data = calgary_fishnet %>% 
                                                as.data.frame() %>%
                                                select(-geometry), 
               method="glm", family="binomial",
               trControl = ctrl)

cvFit

```



```{r cv_hist, warning = FALSE, message = FALSE}
ggplot(as.data.frame(cvFit$resample), aes(Accuracy)) + 
  geom_histogram() +
  scale_x_continuous(limits = c(0, 1)) +
  labs(x="Accuracy",
       y="Count")+
  plotTheme()

```

### 4.3.2. Map predictions

```{r predict_whole, warning = FALSE, message= FALSE}

allPredictions <- 
  predict(cvFit, calgary_fishnet, type="prob")[,2]
  
calgary_fishnet <- 
  cbind(calgary_fishnet,allPredictions) %>%
  mutate(allPredictions = round(allPredictions * 100)) 

```

```{r predict_whole1, warning = FALSE, message= FALSE}

allPredictions <- 
  predict(cvFit, calgary_fishnet, type="prob")[,2]
  
calgary_fishnet <- 
  cbind(calgary_fishnet,allPredictions) %>%
  mutate(allPredictions = round(allPredictions * 100)) 

```

```{r predicted_map1, warning = FALSE, message = FALSE}
# Calculate quantiles and format with 4 decimal places
quantiles_formatted <- sprintf("%.8f", 
                               quantile(calgary_fishnet$allPredictions, 
                                        c(0.1, 0.2, 0.4, 0.6, 0.8), 
                                        na.rm = TRUE))
 # Generate formatted labels manually
quantiles <- quantile(calgary_fishnet$allPredictions.1, 
                      c(0.1, 0.2, 0.4, 0.6, 0.8), 
                      na.rm = TRUE)
quantiles_formatted <- sapply(quantiles, function(x) {
  if (x == 0) {
    "0"
  } else {
    # Extract exponent and mantissa for each quantile
    split_quantile <- unlist(strsplit(format(x, scientific = TRUE), "e"))
    mantissa <- split_quantile[1]
    exponent <- gsub("\\+0|\\+", "", split_quantile[2]) # Remove leading '+0' or '+'
    paste0(mantissa, "*10^", exponent)
  }
})

# Your ggplot code
ggplot() + 
  geom_sf(data = calgary_fishnet, aes(fill = factor(ntile(allPredictions.1, 5))), 
          colour = NA) +
  scale_fill_manual(values = c("#edf8fb", "#b3cde3", "#8c96c6", "#8856a7", "#810f7c"),
                    labels = quantiles_formatted,
                    name = "Predicted\nProbabilities\n(Quintile\nBreaks)") +
  mapTheme() +  # Ensure mapTheme is a defined function that sets theme details
  labs(title = "Title Here")


```

Let’s map it again with the already other land cover types overlaid.

```{r predicted_map2, warning = FALSE, message = FALSE}

 ggplot() + 
  geom_sf(data = calgary_fishnet, aes(fill = factor(ntile(allPredictions.1, 5))), 
          colour = NA) +
  scale_fill_manual(values = c("#edf8fb", "#b3cde3", "#8c96c6", "#8856a7", "#810f7c"),
                    labels = quantiles_formatted,
                    name = "Predicted\nProbabilities\n(Quintile\nBreaks)") +
  geom_sf(data=calgary_fishnet %>% 
              filter(inundation == 1), 
            fill="red",colour=NA) +  
  mapTheme () +
  labs(title="Observed and Predicted Conservation Areas",
       subtitle="Pennsylvania; Existing conserved land in green; Existing development in red ")
```


```{r error_map, warning = FALSE, message= FALSE}
calgary_fishnet %>%
  mutate(confResult=case_when(allPredictions < 50 & inundation==0 ~ "True_Negative",
                              allPredictions >= 50 & inundation==1 ~ "True_Positive",
                              allPredictions < 50 & inundation==1 ~ "False_Negative",
                              allPredictions >= 50 & inundation==0 ~ "False_Positive")) %>%
  ggplot()+
  geom_sf(aes(fill = confResult), color = "transparent")+
  scale_fill_manual(values = c("#edf8fb", "#b3cde3", "#8c96c6", "#8856a7", "#810f7c"),
                    name="Outcomes")+
  labs(title="Confusion Metrics") +
  mapTheme ()

```
